---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# VirtualPals

[![R-CMD-check](https://github.com/your-username/VirtualPals/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/your-username/VirtualPals/actions/workflows/R-CMD-check.yaml)

**VirtualPals** is an R package for creating virtual species and simulating their ecological niches for use in species distribution modeling (SDM) research, algorithm testing, and education. It allows you to define a species' fundamental niche in a multi-dimensional environmental space and project that niche onto a geographic landscape to generate a potential distribution.

The core methodology is based on defining the niche as a multi-dimensional ellipsoid. The environmental conditions inside the ellipsoid are considered "suitable" for the species, while those outside are "unsuitable."

## 🚀 Features

* **Define N-dimensional Niches**: Specify a species' fundamental niche using an ellipsoidal model with a defined center (optimum) and axes (tolerances).
* **Generate Suitability Maps**: Create geographic maps of environmental suitability based on the defined niche and environmental raster data.
* **Create Presence-Absence Data**: Convert continuous suitability maps into binary presence-absence maps.
* **Rich Visualization**:
    * Visualize the environmental niche in a 3D interactive plot.
    * Map the species' potential distribution geographically.
    * Analyze environmental variable correlations.

## 📦 Installation

You can install the development version of VirtualPals from GitHub with:

```{r}
# install.packages("devtools")
# devtools::install_github("your-username/VirtualPals")
```

## quick_start Quick Start Example

Here's a complete workflow to define a virtual species and map its distribution.

### 1. Load Libraries and Data

First, load `VirtualPals` along with `terra` for handling spatial data. You'll need to provide your own environmental raster layers. For this example, we assume you have rasters for Annual Mean Temperature, Annual Precipitation, and Temperature Seasonality.

```{r}
library(VirtualPals)
library(terra)
library(rnaturalearth) # For world map data for masking
# Load environmental raster layers
# Replace with the path to your own data
bio_1 <- terra::rast("path/to/Bio1.tif")
bio_12 <- terra::rast("path/to/Bio12.tif")
bio_4 <- terra::rast("path/to/Bio4.tif")

# Stack and name the layers
bio_stack <- c(bio_1, bio_12, bio_4)
names(bio_stack) <- c("AnnualTemp", "AnnualPrecip", "TempSeasonality")

# Optional: Mask rasters to only include terrestrial areas
world_map_sf <- ne_countries(scale = "medium", returnclass = "sf")
bio_stack_terrestrial <- mask(bio_stack, vect(world_map_sf))
```

### 2. Define the Ecological Niche

Next, define the species' niche. We'll use the `define_niche_ellipsoid()` function (which uses `get_parametric_ellipsoid` internally). You specify the environmental optimum (`center`) and the tolerance along each environmental axis (`axes`).

```{r}
# Define the niche center and axes (breadth)
# Corresponds to: c(AnnualTemp, AnnualPrecip, TempSeasonality)
my_niche <- define_niche_ellipsoid(
  center = c(15, 500, 12000),
  axes = c(10, 500, 2500), # Note: axes are radii (tolerances)
  n_points = 100 # Resolution for the 3D plot surface
)
```

### 3. Generate Suitability Map

Now, project this niche onto the geographic landscape using your environmental data to see where the species could live.

```{r}
# Generate a binary suitability map (1 = suitable, 0 = unsuitable)
suitability_map <- generate_suitability(
  env_rasters = bio_stack_terrestrial,
  niche_definition = my_niche
)
```

### 4. Visualize Results

VirtualPals makes it easy to visualize both the abstract niche in environmental space and the final species distribution.

#### Plot the Geographic Distribution

```{r}
# Plot the binary distribution map
plot(suitability_map,
     col = c("grey90", "darkgreen"),
     legend = FALSE,
     main = "Potential Distribution of Virtualis PALSicus")

# Add country borders for context
lines(vect(ne_countries(scale = "medium", returnclass = "sf")), col = "black", lwd = 0.5)
```


#### Plot the Niche in 3D Environmental Space

Use the `plot_niche_3d()` function for an interactive visualization of the environmental space.

```{r}
# This function requires the 'plotly' package
# It takes the environmental data and the niche object
plot_niche_3d(
  env_data = bio_stack_terrestrial,
  niche_definition = my_niche
)
```
